% -*- coding: utf-8 -*-
\documentclass[letterpaper]{book}

\input{preamble}
\setcounter{chapter}{2}

\begin{document}

%\chapter{Characters}\label{char}
\chapter{Characters}\label{char}

%Internally, \TeX\ represents characters by their (integer) 
%character code. This chapter treats those codes, and the
%commands that have access to them.
Internally, \TeX\ represents characters by their (integer) 
character code. This chapter treats those codes, and the
commands that have access to them.

%\label{cschap:char}\label{cschap:chardef}\label{cschap:accent}\label{cschap:uccode}\label{cschap:lccode}\label{cschap:uppercase}\label{cschap:lowercase}\label{cschap:string}\label{cschap:escapechar}
%\begin{inventory}
%\item [\cs{char}]
%      Explicit denotation of a character to be typeset. 
\label{cschap:char}\label{cschap:chardef}\label{cschap:accent}\label{cschap:uccode}\label{cschap:lccode}\label{cschap:uppercase}\label{cschap:lowercase}\label{cschap:string}\label{cschap:escapechar}
\begin{inventory}
\item [\cs{char}]
      Explicit denotation of a character to be typeset. 

%\item [\cs{chardef}] 
%      Define a control sequence to be a synonym for
%      a~character code.
\item [\cs{chardef}] 
      Define a control sequence to be a synonym for
      a~character code.

%\item [\cs{accent}] 
%      Command to place accent characters.
\item [\cs{accent}] 
      Command to place accent characters.

%\item [\cs{if}]
%      Test equality of character codes. 
\item [\cs{if}]
      Test equality of character codes. 

%\item [\cs{ifx}]
%      Test equality of both character and category codes.
\item [\cs{ifx}]
      Test equality of both character and category codes.

%\item [\cs{let}]
%      Define a control sequence to be a synonym of a token.
\item [\cs{let}]
      Define a control sequence to be a synonym of a token.

%\item [\cs{uccode}] 
%      Query or set
%      the character code that is the uppercase variant of a given code.
\item [\cs{uccode}] 
      Query or set
      the character code that is the uppercase variant of a given code.

%\item [\cs{lccode}]
%      Query or set
%      the character code that is the lowercase variant of a given code.
\item [\cs{lccode}]
      Query or set
      the character code that is the lowercase variant of a given code.

%\item [\cs{uppercase}]
%      Convert the \gr{general text} argument to its uppercase form.
\item [\cs{uppercase}]
      Convert the \gr{general text} argument to its uppercase form.

%\item [\cs{lowercase}] 
%      Convert the \gr{general text} argument to its lowercase form.
\item [\cs{lowercase}] 
      Convert the \gr{general text} argument to its lowercase form.

%\item [\cs{string}]
%      Convert a token to a string of one or more characters.
%\item [\cs{escapechar}]
%      Number of the character that is to be used 
%      for the escape character
%      when control sequences are being converted
%      into character tokens. \IniTeX\ default:~92~(\cs{}).
%\end{inventory}
\item [\cs{string}]
      Convert a token to a string of one or more characters.
\item [\cs{escapechar}]
      Number of the character that is to be used 
      for the escape character
      when control sequences are being converted
      into character tokens. \IniTeX\ default:~92~(\cs{}).
\end{inventory}

%%\point[char:code] Character codes
%\section{Character codes}
%\label{char:code}
%\point[char:code] Character codes
\section{Character codes}
\label{char:code}

%Conceptually it is easiest to think that \TeX\ works with
%characters internally, but in fact
%\TeX\ works with integers: the \indextermsub{character}{codes}. 
Conceptually it is easiest to think that \TeX\ works with
characters internally, but in fact
\TeX\ works with integers: the \indextermsub{character}{codes}. 

%The way characters are encoded in a computer may differ
%from system to system.
%Therefore \TeX\ uses its own scheme of character codes.
%Any character that is read from a file (or from the user terminal)
%is converted to a character code according to the
%character code table.
%A~category code is then assigned based on this (see Chapter~\ref{mouth}).
%The character code table is based on the 7-bit \ascii{} table
%for numbers under~128 (see Section~\ref{sec:asciitable}).
The way characters are encoded in a computer may differ
from system to system.
Therefore \TeX\ uses its own scheme of character codes.
Any character that is read from a file (or from the user terminal)
is converted to a character code according to the
character code table.
A~category code is then assigned based on this (see Chapter~\ref{mouth}).
The character code table is based on the 7-bit \ascii{} table
for numbers under~128 (see Section~\ref{sec:asciitable}).

%There is an explicit conversion between characters
%(better:  character tokens)
%and  character codes  using the left quote (grave, back quote)
%character~\n{`{}}:
%at all places where \TeX\ expects a \gram{number} you
%can use the left quote followed by a character
%token or
%a single-character control sequence.
%Thus both \verb.\count`a. and \verb.\count`\a. are synonyms
%for \verb.\count97.. See also Chapter~\ref{number}.
There is an explicit conversion between characters
(better:  character tokens)
and  character codes  using the left quote (grave, back quote)
character~\n{`{}}:
at all places where \TeX\ expects a \gram{number} you
can use the left quote followed by a character
token or
a single-character control sequence.
Thus both \verb.\count`a. and \verb.\count`\a. are synonyms
for \verb.\count97.. See also Chapter~\ref{number}.

%The possibility of a single-character control
%sequence is necessary in certain cases such as
%\begin{disp}\verb>\catcode`\%=11>\quad or\quad \verb>\def\CommentSign{\char`\%}>\end{disp}
%which would be misunderstood if the backslash were left out.
%For instance
%\begin{verbatim}
%\catcode`%=11
%\end{verbatim}
%would consider
%the \n{=11} to be a comment.
%Single-character
%control sequences can be formed from characters with any
%category code.
The possibility of a single-character control
sequence is necessary in certain cases such as
\begin{disp}\verb>\catcode`\%=11>\quad or\quad \verb>\def\CommentSign{\char`\%}>\end{disp}
which would be misunderstood if the backslash were left out.
For instance
\begin{verbatim}
\catcode`%=11
\end{verbatim}
would consider
the \n{=11} to be a comment.
Single-character
control sequences can be formed from characters with any
category code.

%After the conversion to character codes any connection
%with external representations has disappeared. Of course,
%for most characters  the visible output will `equal' the input
%(that is, an `\n{a}' causes an~`a').
%There are exceptions, however, even among the common symbols.
%In the Computer Modern
%roman fonts there are no `less than' and `greater than'
%\message{Check <>! Dammit!}%
%signs, so the input `\verb.<>.' will give `<>' in the output.
%%{\MathRMx<>}
After the conversion to character codes any connection
with external representations has disappeared. Of course,
for most characters  the visible output will `equal' the input
(that is, an `\n{a}' causes an~`a').
There are exceptions, however, even among the common symbols.
In the Computer Modern
roman fonts there are no `less than' and `greater than'
\message{Check <>! Dammit!}%
signs, so the input `\verb.<>.' will give `<>' in the output.
%{\MathRMx<>}

%In order to make \TeX\ machine independent at the output
%side, the character codes are also used in the \n{dvi} file:
%opcodes $n=0\ldots127$ denote simply the instruction `take
%character $n$ from the current font'. The complete definition
%of the opcodes in a \n{dvi} file can be found in~\cite{Knuth:TeXprogram}.
In order to make \TeX\ machine independent at the output
side, the character codes are also used in the \n{dvi} file:
opcodes $n=0\ldots127$ denote simply the instruction `take
character $n$ from the current font'. The complete definition
of the opcodes in a \n{dvi} file can be found in~\cite{Knuth:TeXprogram}.


%%\point Control sequences for characters
%\section{Control sequences for characters}
%\point Control sequences for characters
\section{Control sequences for characters}

%There are a number of ways in which a control sequence can denote
%a character. The \cs{char} command specifies a character to be
%typeset; the \cs{let} command introduces
%a synonym for a character token, that is,
%the combination of character code and category code.
There are a number of ways in which a control sequence can denote
a character. The \cs{char} command specifies a character to be
typeset; the \cs{let} command introduces
a synonym for a character token, that is,
the combination of character code and category code.

%%\point Denoting characters to be typeset: \cs\char
%\section{Denoting characters to be typeset: \protect\cs{char}}
%\point Denoting characters to be typeset: \cs\char
\section{Denoting characters to be typeset: \protect\cs{char}}

%Characters can be denoted numerically by, for example,
%\verb.\char98.\cstoidx char\par.
%This command tells \TeX\ to add character number~98 of the
%current font to the horizontal list currently under construction.
Characters can be denoted numerically by, for example,
\verb.\char98.\cstoidx char\par.
This command tells \TeX\ to add character number~98 of the
current font to the horizontal list currently under construction.

%Instead of decimal notation, it is often more convenient to
%use octal or hexadecimal notation. For octal the single quote is used:
%\verb.\char'142.; hexadecimal uses the double quote: \verb.\char"62..
%Note that \verb.\char''62. is incorrect; the process that replaces
%two quotes by a double quote works at a later stage of processing
%(the visual processor) than number scanning (the execution processor).
Instead of decimal notation, it is often more convenient to
use octal or hexadecimal notation. For octal the single quote is used:
\verb.\char'142.; hexadecimal uses the double quote: \verb.\char"62..
Note that \verb.\char''62. is incorrect; the process that replaces
two quotes by a double quote works at a later stage of processing
(the visual processor) than number scanning (the execution processor).

%Because of the explicit conversion to character codes by the
%back quote character it is also possible to get a `b' \ldash provided
%that you are using a font organized a bit like the \ascii{} table \rdash
%with \verb.\char`b.  or \verb.\char`\b..
Because of the explicit conversion to character codes by the
back quote character it is also possible to get a `b' \ldash provided
that you are using a font organized a bit like the \ascii{} table \rdash
with \verb.\char`b.  or \verb.\char`\b..

%The \cs{char} command looks superficially a bit like
%the \verb-^^- substitution mechanism (Chapter~\ref{mouth}).
%Both mechanisms access characters without directly denoting them.
%However, the \verb-^^- mechanism operates in a very early stage of
%processing (in the input processor of \TeX,
%but before category code
%assignment); the \cs{char} command, on the other hand,
%comes in the final stages of processing. 
%In effect it says `typeset character number
%so-and-so'.
The \cs{char} command looks superficially a bit like
the \verb-^^- substitution mechanism (Chapter~\ref{mouth}).
Both mechanisms access characters without directly denoting them.
However, the \verb-^^- mechanism operates in a very early stage of
processing (in the input processor of \TeX,
but before category code
assignment); the \cs{char} command, on the other hand,
comes in the final stages of processing. 
In effect it says `typeset character number
so-and-so'.

%There is a construction to let a control sequence stand
%for some character code: the \csterm chardef\par\ command.
%The syntax of this is \label{chardef}
%\begin{disp}\cs{chardef}\gram{control sequence}\gr{equals}\gram{number}, 
%\end{disp}
%where the number can be an explicit
%representation or a counter value, but it can also be
%a character code
%obtained using the left quote command (see above; 
%the full definition of \gr{number} is given in Chapter~\ref{number}). 
%In the plain format 
%the latter possibility is used in
%definitions such as
%\begin{verbatim}
%\chardef\%=`\%
%\end{verbatim}
%which could have been given equivalently as
%\begin{verbatim}
%\chardef\%=37
%\end{verbatim}
%After this command, the control symbol \verb>\%>
%used on its own is a synonym for \verb>\char37>,
%that is, the command to typeset character~37
%(usually the per cent character).
There is a construction to let a control sequence stand
for some character code: the \csterm chardef\par\ command.
The syntax of this is \label{chardef}
\begin{disp}\cs{chardef}\gram{control sequence}\gr{equals}\gram{number}, 
\end{disp}
where the number can be an explicit
representation or a counter value, but it can also be
a character code
obtained using the left quote command (see above; 
the full definition of \gr{number} is given in Chapter~\ref{number}). 
In the plain format 
the latter possibility is used in
definitions such as
\begin{verbatim}
\chardef\%=`\%
\end{verbatim}
which could have been given equivalently as
\begin{verbatim}
\chardef\%=37
\end{verbatim}
After this command, the control symbol \verb>\%>
used on its own is a synonym for \verb>\char37>,
that is, the command to typeset character~37
(usually the per cent character).

%A control sequence that has been defined with a \cs{chardef}
%command can also be used as a \gr{number}.
%This fact is used in  allocation commands such as 
%\cs{newbox} (see Chapters~\ref{number} and~\ref{alloc}).
%Tokens defined with \cs{mathchardef} can also be used this
%way.
A control sequence that has been defined with a \cs{chardef}
command can also be used as a \gr{number}.
This fact is used in  allocation commands such as 
\cs{newbox} (see Chapters~\ref{number} and~\ref{alloc}).
Tokens defined with \cs{mathchardef} can also be used this
way.

%\subsection{Implicit character tokens: \protect\cs{let}}
\subsection{Implicit character tokens: \protect\cs{let}}

%Another construction defining a control sequence
%to stand for (among other things)
%a character is~\cs{let}\cstoidx let\par:
%\begin{disp}\cs{let}\gr{control sequence}\gr{equals}\gr{token}\end{disp}
%with a character token on the right hand side of the (optional)
%equals sign. The result is called an \indextermbus{implicit}{character} token.
%(See page~\pageref{let} for a further discussion of~\cs{let}.)
Another construction defining a control sequence
to stand for (among other things)
a character is~\cs{let}\cstoidx let\par:
\begin{disp}\cs{let}\gr{control sequence}\gr{equals}\gr{token}\end{disp}
with a character token on the right hand side of the (optional)
equals sign. The result is called an \indextermbus{implicit}{character} token.
(See page~\pageref{let} for a further discussion of~\cs{let}.)

%In the
%plain format there are for instance synonyms for
%the open and close brace:
%\begin{verbatim}
%\let\bgroup={ \let\egroup=}
%\end{verbatim}
%The resulting control sequences are called `implicit braces'
%(see Chapter~\ref{group}).
In the
plain format there are for instance synonyms for
the open and close brace:
\begin{verbatim}
\let\bgroup={ \let\egroup=}
\end{verbatim}
The resulting control sequences are called `implicit braces'
(see Chapter~\ref{group}).

%Assigning characters by \cs{let}
%is different from defining control sequences by \cs{chardef}, 
%in the sense that \cs{let}
%makes the control sequence stand for the combination
%of a character code and category code. 
Assigning characters by \cs{let}
is different from defining control sequences by \cs{chardef}, 
in the sense that \cs{let}
makes the control sequence stand for the combination
of a character code and category code. 

%As an example
%\begin{verbatim}
%\catcode`|=2 % make the bar an end of group
%\let\b=|  % make \b a bar character
%{\def\m{...}\b \m
%\end{verbatim}
%gives an `undefined control sequence \cs{m}'
%because the \cs{b} closed the group inside which \cs{m}
%was defined. On the other hand,
%\begin{verbatim}
%\let\b=| % make \b a bar character
%\catcode`|=2  % make the bar character end of group
%{\def\m{...}\b \m
%\end{verbatim}
%leaves one group open, and it prints a vertical bar
%(or whatever is in position 124 of the current font).
%The first of these examples
%implies that even when the braces have been redefined
%(for instance into active characters for macros that
%format C code) the beginning-of-group and end-of-group
%functionality is available through the control sequences
%\cs{bgroup} and~\cs{egroup}.
As an example
\begin{verbatim}
\catcode`|=2 % make the bar an end of group
\let\b=|  % make \b a bar character
{\def\m{...}\b \m
\end{verbatim}
gives an `undefined control sequence \cs{m}'
because the \cs{b} closed the group inside which \cs{m}
was defined. On the other hand,
\begin{verbatim}
\let\b=| % make \b a bar character
\catcode`|=2  % make the bar character end of group
{\def\m{...}\b \m
\end{verbatim}
leaves one group open, and it prints a vertical bar
(or whatever is in position 124 of the current font).
The first of these examples
implies that even when the braces have been redefined
(for instance into active characters for macros that
format C code) the beginning-of-group and end-of-group
functionality is available through the control sequences
\cs{bgroup} and~\cs{egroup}.

%Here is
%another example to show
%that implicit character tokens are hard to distinguish
%from real character tokens. After the above sequence
%\begin{verbatim}
%\catcode`|=2 \let\b=|
%\end{verbatim}
%the tests
%\begin{verbatim}
%\if\b|
%\end{verbatim}
%and
%\begin{verbatim}
%\ifcat\b}
%\end{verbatim}
%are both true.
Here is
another example to show
that implicit character tokens are hard to distinguish
from real character tokens. After the above sequence
\begin{verbatim}
\catcode`|=2 \let\b=|
\end{verbatim}
the tests
\begin{verbatim}
\if\b|
\end{verbatim}
and
\begin{verbatim}
\ifcat\b}
\end{verbatim}
are both true.

%Yet another example can be found in the plain format:
%the commands
%\begin{verbatim}
%\let\sp=^ \let\sb=_ 
%\end{verbatim}
%allow people without an
%underscore or circumflex on their keyboard to 
%make sub- and superscripts in mathematics.
%For instance:
%\begin{disp}\verb>x\sp2\sb{ij}>\quad gives\quad $x\sp2\sb{ij}$\end{disp}
%If a person typing in the format itself does not have
%these keys, some further tricks are needed:\label{spsb:truc}
%\begin{verbatim}
%{\lccode`,=94 \lccode`.=95 \catcode`,=7 \catcode`.=8
%\lowercase{\global\let\sp=, \global\let\sb=.}}
%\end{verbatim}
%will do the job; see below for an explanation of lowercase codes.
%The \verb>^^> method as it was in \TeX\ version~2
%(see page~\pageref{hathat}) cannot be used here,
%as it would require typing two characters that can ordinarily
%not be input.
%With the extension in \TeX\ version~3 it would also be possible
%to write
%\begin{verbatim}
%{\catcode`\,=7
%\global\let\sp=,,5e \global\let\sb=,,5f}
%\end{verbatim}
%denoting the codes 94 and 95 hexadecimally.
Yet another example can be found in the plain format:
the commands
\begin{verbatim}
\let\sp=^ \let\sb=_ 
\end{verbatim}
allow people without an
underscore or circumflex on their keyboard to 
make sub- and superscripts in mathematics.
For instance:
\begin{disp}\verb>x\sp2\sb{ij}>\quad gives\quad $x\sp2\sb{ij}$\end{disp}
If a person typing in the format itself does not have
these keys, some further tricks are needed:\label{spsb:truc}
\begin{verbatim}
{\lccode`,=94 \lccode`.=95 \catcode`,=7 \catcode`.=8
\lowercase{\global\let\sp=, \global\let\sb=.}}
\end{verbatim}
will do the job; see below for an explanation of lowercase codes.
The \verb>^^> method as it was in \TeX\ version~2
(see page~\pageref{hathat}) cannot be used here,
as it would require typing two characters that can ordinarily
not be input.
With the extension in \TeX\ version~3 it would also be possible
to write
\begin{verbatim}
{\catcode`\,=7
\global\let\sp=,,5e \global\let\sb=,,5f}
\end{verbatim}
denoting the codes 94 and 95 hexadecimally.

%Finding out just what a control sequence has been defined to be with
%\cs{let} can be done using \cs{meaning}:
%the sequence
%\begin{verbatim}
%\let\x=3 \meaning\x
%\end{verbatim}
%gives
%`\n{the character 3}'.
Finding out just what a control sequence has been defined to be with
\cs{let} can be done using \cs{meaning}:
the sequence
\begin{verbatim}
\let\x=3 \meaning\x
\end{verbatim}
gives
`\n{the character 3}'.

%%\point Accents
%\section{Accents}
%\point Accents
\section{Accents}

%\emph{Accents}\index{accents} can be placed by the
%\gr{horizontal command}~\csterm accent\par
%\label{character}:
%\begin{disp}\cs{accent}\gr{8-bit number}\gr{optional assignments}%
%     \gr{character}\end{disp} 
%where \gr{character} is a character of
%category 11\index{category!11} or~12\index{category!12},
%a~\cs{char}\gr{8-bit number} command, or a~\cs{chardef} token. If none
%of these four types of \gr{character} follows, the accent is taken to
%be a \cs{char} command itself; this gives an accent `suspended in
%mid-air'. Otherwise the accent is placed on top of the following
%character.  Font changes between the accent and the character can be
%effected by the \gr{optional assignments}.
\emph{Accents}\index{accents} can be placed by the
\gr{horizontal command}~\csterm accent\par
\label{character}:
\begin{disp}\cs{accent}\gr{8-bit number}\gr{optional assignments}%
     \gr{character}\end{disp} 
where \gr{character} is a character of
category 11\index{category!11} or~12\index{category!12},
a~\cs{char}\gr{8-bit number} command, or a~\cs{chardef} token. If none
of these four types of \gr{character} follows, the accent is taken to
be a \cs{char} command itself; this gives an accent `suspended in
mid-air'. Otherwise the accent is placed on top of the following
character.  Font changes between the accent and the character can be
effected by the \gr{optional assignments}.

%An unpleasant implication of the fact that an \cs{accent} command
%has to be followed by a \gr{character} is that it is not
%possible to place an accent on a ligature, or
%two accents on top of each other.
%In some languages, such as Hindi or Vietnamese,
%such double accents do occur.
%Positioning accents on top of each other is possible,
%however, in math mode.
An unpleasant implication of the fact that an \cs{accent} command
has to be followed by a \gr{character} is that it is not
possible to place an accent on a ligature, or
two accents on top of each other.
In some languages, such as Hindi or Vietnamese,
such double accents do occur.
Positioning accents on top of each other is possible,
however, in math mode.

%The width of a character with an accent is the same as that of
%the unaccented character. \TeX\ assumes that the 
%accent as it appears in the font file
%is properly positioned for a character that is as high
%as the x-height of the font; for characters with other heights
%it correspondingly lowers or raises the accent.
The width of a character with an accent is the same as that of
the unaccented character. \TeX\ assumes that the 
accent as it appears in the font file
is properly positioned for a character that is as high
as the x-height of the font; for characters with other heights
it correspondingly lowers or raises the accent.

%No genuine under-accents exist in \TeX. They are
%implemented as low placed over-accents. A~way of handling
%them more correctly would be to write a macro that
%measures the following character, and raises or drops
%the accent accordingly.
%The cedilla macro, \cs{c}\cstoidx c\par,
%in plain \TeX\ does something along these lines. However,
%it does not drop the accent for characters with descenders.
No genuine under-accents exist in \TeX. They are
implemented as low placed over-accents. A~way of handling
them more correctly would be to write a macro that
measures the following character, and raises or drops
the accent accordingly.
The cedilla macro, \cs{c}\cstoidx c\par,
in plain \TeX\ does something along these lines. However,
it does not drop the accent for characters with descenders.

%The horizontal positioning of an accent is controlled by
%\cs{fontdimen1}, \indextermsub{slant}{per point}. Kerns are used
%for the horizontal movement. Note that, although they
%are inserted automatically, these kerns are classified
%as {\italic explicit\/} kerns. Therefore they inhibit hyphenation
%in the parts of the word before and after the kern.
The horizontal positioning of an accent is controlled by
\cs{fontdimen1}, \indextermsub{slant}{per point}. Kerns are used
for the horizontal movement. Note that, although they
are inserted automatically, these kerns are classified
as {\italic explicit\/} kerns. Therefore they inhibit hyphenation
in the parts of the word before and after the kern.

%As an example of kerning for accents, 
%here follows the dump of a horizontal list.
%\message{maybe italic correction for extra line}
%\begin{verbatim}
%\setbox0=\hbox{\it \`l}
%\showbox0
%\end{verbatim}
%gives
%\begin{verbatim}
%\hbox(9.58334+0.0)x2.55554
%.\kern -0.61803 (for accent)
%.\hbox(6.94444+0.0)x5.11108, shifted -2.6389
%..\tenit ^^R
%.\kern -4.49306 (for accent)
%.\tenit l
%\end{verbatim}
%Note that the accent is placed first, so afterwards the italic
%correction of the last character is still available.
As an example of kerning for accents, 
here follows the dump of a horizontal list.
\message{maybe italic correction for extra line}
\begin{verbatim}
\setbox0=\hbox{\it \`l}
\showbox0
\end{verbatim}
gives
\begin{verbatim}
\hbox(9.58334+0.0)x2.55554
.\kern -0.61803 (for accent)
.\hbox(6.94444+0.0)x5.11108, shifted -2.6389
..\tenit ^^R
.\kern -4.49306 (for accent)
.\tenit l
\end{verbatim}
Note that the accent is placed first, so afterwards the italic
correction of the last character is still available.

%\section{Testing characters}
\section{Testing characters}

%Equality of character codes is tested by \cs{if}:
%\begin{disp}\cs{if}\gr{token$_1$}\gr{token$_2$}\end{disp}
%Tokens following this conditional are expanded until two
%unexpandable tokens are left. The condition is then true
%if those tokens are character tokens with the same character
%code, regardless of category code. 
Equality of character codes is tested by \cs{if}:
\begin{disp}\cs{if}\gr{token$_1$}\gr{token$_2$}\end{disp}
Tokens following this conditional are expanded until two
unexpandable tokens are left. The condition is then true
if those tokens are character tokens with the same character
code, regardless of category code. 

%An unexpandable control
%sequence is considered to have character code 256 and
%category code~16\index{category!16}
%(so that it is unequal to anything except
%another control sequence), except in the case
%where it had been \cs{let} to a non-active character token.
%In that case it is considered to have the character code
%and category code of that character. This was mentioned above.
An unexpandable control
sequence is considered to have character code 256 and
category code~16\index{category!16}
(so that it is unequal to anything except
another control sequence), except in the case
where it had been \cs{let} to a non-active character token.
In that case it is considered to have the character code
and category code of that character. This was mentioned above.

%The test \cs{ifcat} for category codes was mentioned
%in Chapter~\ref{mouth}; the test
%\begin{disp}\cs{ifx}\gr{token$_1$}\gr{token$_2$}\end{disp}
%can be used to test for category code and character code
%simultaneously.
%The tokens following this test are not expanded.
%However, if they are macros, \TeX\
%tests their expansions for equality.
The test \cs{ifcat} for category codes was mentioned
in Chapter~\ref{mouth}; the test
\begin{disp}\cs{ifx}\gr{token$_1$}\gr{token$_2$}\end{disp}
can be used to test for category code and character code
simultaneously.
The tokens following this test are not expanded.
However, if they are macros, \TeX\
tests their expansions for equality.

%Quantities defined by \cs{chardef} can be tested with
%\cs{ifnum}:
%\begin{verbatim}
%\chardef\a=`x \chardef\b=`y \ifnum\a=\b % is false 
%\end{verbatim}
%based on the fact (see Chapter~\ref{number}) that
%\gr{chardef token}s can be used as numbers.
Quantities defined by \cs{chardef} can be tested with
\cs{ifnum}:
\begin{verbatim}
\chardef\a=`x \chardef\b=`y \ifnum\a=\b % is false 
\end{verbatim}
based on the fact (see Chapter~\ref{number}) that
\gr{chardef token}s can be used as numbers.

%See also section~\ref{sec:charactertests}
See also section~\ref{sec:charactertests}


%\section{Uppercase and lowercase}
\section{Uppercase and lowercase}

%%\spoint[uc/lc] Uppercase and lowercase codes
%\subsection{Uppercase and lowercase codes}
%\label{uc/lc}
%\spoint[uc/lc] Uppercase and lowercase codes
\subsection{Uppercase and lowercase codes}
\label{uc/lc}

%To each of the character codes correspond\cstoidx lccode\par\cstoidx uccode\par
%an \indextermsub{uppercase}{code}\index{code!uppercase|see{uppercase, code}}
%and a \indextermsub{lowercase}{code}\index{code!lowercase|see{lowercase, code}}
%(for still more codes see below).
%These can be assigned
%by 
%\begin{Disp}\cs{uccode}\gram{number}\gr{equals}\gram{number}\end{Disp}
%and 
%\begin{Disp}\cs{lccode}\gram{number}\gr{equals}\gram{number}.\end{Disp}
%In \IniTeX\ codes \verb-`a..`z-, \verb-`A..`Z- have uppercase code
%\label{ini:uclc}
%\verb-`A..`Z- and lowercase code \verb-`a..`z-.
%All other character codes have both uppercase and lowercase
%code zero.
To each of the character codes correspond\cstoidx lccode\par\cstoidx uccode\par
an \indextermsub{uppercase}{code}\index{code!uppercase|see{uppercase, code}}
and a \indextermsub{lowercase}{code}\index{code!lowercase|see{lowercase, code}}
(for still more codes see below).
These can be assigned
by 
\begin{Disp}\cs{uccode}\gram{number}\gr{equals}\gram{number}\end{Disp}
and 
\begin{Disp}\cs{lccode}\gram{number}\gr{equals}\gram{number}.\end{Disp}
In \IniTeX\ codes \verb-`a..`z-, \verb-`A..`Z- have uppercase code
\label{ini:uclc}
\verb-`A..`Z- and lowercase code \verb-`a..`z-.
All other character codes have both uppercase and lowercase
code zero.

%%\spoint[upcase] Uppercase and lowercase commands
%\subsection{Uppercase and lowercase commands}
%\label{upcase}
%\spoint[upcase] Uppercase and lowercase commands
\subsection{Uppercase and lowercase commands}
\label{upcase}

%The commands \verb-\uppercase{...}- and \verb-\lowercase{...}-
%\cstoidx uppercase\par\cstoidx lowercase\par
%go through their argument lists, replacing all character 
%codes of explicit character tokens
%by their uppercase and lowercase code respectively
%if these are non-zero,
%without changing the category codes. 
The commands \verb-\uppercase{...}- and \verb-\lowercase{...}-
\cstoidx uppercase\par\cstoidx lowercase\par
go through their argument lists, replacing all character 
codes of explicit character tokens
by their uppercase and lowercase code respectively
if these are non-zero,
without changing the category codes. 

%The argument of \cs{uppercase} and \cs{lowercase}
%is a \gr{general text}, which is defined as
%\begin{Disp} \gr{general text} $\longrightarrow$ \gr{filler}\lb
%      \gr{balanced text}\gr{right brace}\end{Disp}
%(for the definition of \gr{filler} see Chapter~\ref{gramm})
%meaning that the left brace can be implicit, but the closing
%right brace must be an explicit character token with category
%code~2. \TeX\ performs expansion to find the opening
%brace.
The argument of \cs{uppercase} and \cs{lowercase}
is a \gr{general text}, which is defined as
\begin{Disp} \gr{general text} $\longrightarrow$ \gr{filler}\lb
      \gr{balanced text}\gr{right brace}\end{Disp}
(for the definition of \gr{filler} see Chapter~\ref{gramm})
meaning that the left brace can be implicit, but the closing
right brace must be an explicit character token with category
code~2. \TeX\ performs expansion to find the opening
brace.

%Uppercasing and lowercasing are executed in the execution processor;
%they are not `macro expansion' activities
%like \cs{number} or \cs{string}.
%The sequence (attempting to produce~\cs{A})
%\begin{verbatim}
%\expandafter\csname\uppercase{a}\endcsname
%\end{verbatim}
%gives an error (\TeX\ inserts an \cs{endcsname} before   the
%\cs{uppercase} because \cs{uppercase} is unexpandable), but
%\begin{verbatim}
%\uppercase{\csname a\endcsname}
%\end{verbatim}
%works.
Uppercasing and lowercasing are executed in the execution processor;
they are not `macro expansion' activities
like \cs{number} or \cs{string}.
The sequence (attempting to produce~\cs{A})
\begin{verbatim}
\expandafter\csname\uppercase{a}\endcsname
\end{verbatim}
gives an error (\TeX\ inserts an \cs{endcsname} before   the
\cs{uppercase} because \cs{uppercase} is unexpandable), but
\begin{verbatim}
\uppercase{\csname a\endcsname}
\end{verbatim}
works.

%As an example of the correct use of \cs{uppercase}, here
%is a macro that tests if a character is uppercase:
%\begin{verbatim}
%\def\ifIsUppercase#1{\uppercase{\if#1}#1}
%\end{verbatim}
%The same test can be
%performed by \verb>\ifnum`#1=\uccode`#1>.
As an example of the correct use of \cs{uppercase}, here
is a macro that tests if a character is uppercase:
\begin{verbatim}
\def\ifIsUppercase#1{\uppercase{\if#1}#1}
\end{verbatim}
The same test can be
performed by \verb>\ifnum`#1=\uccode`#1>.

%Hyphenation of words starting with an uppercase character,
%that is, a character not equal to its own \cs{lccode},
%is subject to the \cs{uchyph} parameter: if this
%is positive, hyphenation of capitalized words is allowed.
%See also Chapter~\ref{line:break}.
Hyphenation of words starting with an uppercase character,
that is, a character not equal to its own \cs{lccode},
is subject to the \cs{uchyph} parameter: if this
is positive, hyphenation of capitalized words is allowed.
See also Chapter~\ref{line:break}.

%%\spoint Uppercase and lowercase forms of keywords
%\subsection{Uppercase and lowercase forms of keywords}
%\spoint Uppercase and lowercase forms of keywords
\subsection{Uppercase and lowercase forms of keywords}

%Each character in \TeX\ keywords, such as \n{pt}, can be
%given in uppercase or lowercase form. 
%For instance, \n{pT}, \n{Pt}, \n{pt}, and~\n{PT} all have
%the same meaning. \TeX\ does not use
%the \cs{uccode} and \cs{lccode} tables here to
%determine the lowercase form. Instead it
%converts uppercase characters to lowercase by adding~32
%\ldash the \ascii{} difference between uppercase and lowercase
%characters \rdash to their character code. This has some implications
%for implementations of \TeX\ for non-roman alphabets;
%see page 370 of \TeXbook, \cite{Knuth:TeXbook}.
Each character in \TeX\ keywords, such as \n{pt}, can be
given in uppercase or lowercase form. 
For instance, \n{pT}, \n{Pt}, \n{pt}, and~\n{PT} all have
the same meaning. \TeX\ does not use
the \cs{uccode} and \cs{lccode} tables here to
determine the lowercase form. Instead it
converts uppercase characters to lowercase by adding~32
\ldash the \ascii{} difference between uppercase and lowercase
characters \rdash to their character code. This has some implications
for implementations of \TeX\ for non-roman alphabets;
see page 370 of \TeXbook, \cite{Knuth:TeXbook}.

%\subsection{Creative use of \cs{uppercase} and \cs{lowercase}}
\subsection{Creative use of \cs{uppercase} and \cs{lowercase}}

%The fact that \cs{uppercase} and \cs{lowercase} do not change
%category codes can sometimes be used to create certain
%character-code--category-code combinations that would
%otherwise be difficult to produce. See for instance the
%explanation of the \cs{newif} macro in Chapter~\ref{if},
%and another example on page~\pageref{spsb:truc}.
The fact that \cs{uppercase} and \cs{lowercase} do not change
category codes can sometimes be used to create certain
character-code--category-code combinations that would
otherwise be difficult to produce. See for instance the
explanation of the \cs{newif} macro in Chapter~\ref{if},
and another example on page~\pageref{spsb:truc}.

%For a slightly different application, consider the
%problem (solved by Rainer Sch\"opf) of,
%given a counter \verb-\newcount\mycount-, writing character
%number \verb-\mycount- to the terminal.
%Here is a solution:
%%\begin{verbatim}
%%\lccode`a=\mycount \chardef\terminal=16
%%\lowercase{\write\terminal{a}}
%%\end{verbatim}
%\begin{verbatim}
%\lccode`a=\mycount \chardef\terminal=16
%\end{verbatim}
%\begin{verbatim}
%\lowercase{\write\terminal{a}}
%\end{verbatim}
%The \cs{lowercase} command effectively changes the 
%argument of the \cs{write} command from~`\n a'
%into whatever it should be.
For a slightly different application, consider the
problem (solved by Rainer Sch\"opf) of,
given a counter \verb-\newcount\mycount-, writing character
number \verb-\mycount- to the terminal.
Here is a solution:
%\begin{verbatim}
%\lccode`a=\mycount \chardef\terminal=16
%\lowercase{\write\terminal{a}}
%\end{verbatim}
\begin{verbatim}
\lccode`a=\mycount \chardef\terminal=16
\end{verbatim}
\begin{verbatim}
\lowercase{\write\terminal{a}}
\end{verbatim}
The \cs{lowercase} command effectively changes the 
argument of the \cs{write} command from~`\n a'
into whatever it should be.

%%\point[codename] Codes of a character
%\section{Codes of a character}
%\label{codename}
%\point[codename] Codes of a character
\section{Codes of a character}
\label{codename}

%Each character code has a number of \gr{codename}s 
%associated\indexterm{codenames}
%with it. These are integers in various ranges that determine
%how the character is treated in various contexts, or
%how the occurrence of that character changes the workings
%of \TeX\ in certain contexts.
Each character code has a number of \gr{codename}s 
associated\indexterm{codenames}
with it. These are integers in various ranges that determine
how the character is treated in various contexts, or
how the occurrence of that character changes the workings
of \TeX\ in certain contexts.

%The code names are as follows:
%\begin{description}\item [\cs{catcode}]
%\gr{4-bit number} (0--15); the category to which a character belongs.
%This is treated in Chapter~\ref{mouth}.
%\item [\cs{mathcode}]
%\gr{15-bit number} (0--\verb-"7FFF-) or \verb-"8000-;
%determines how a character is treated
%in math mode. See Chapter~\ref{mathchar}.
%\item [\cs{delcode}]
%\gr{27-bit number} (0--\n{\hex7$\,$FFF$\,$FFF});
%determines how a character is treated after
%\cs{left} or \cs{right} in math mode.
%See page~\pageref{delcodes}.
%\item [\cs{sfcode}]
%integer; determines how spacing is affected after this character.
%See Chapter~\ref{space}.
%\item [\cs{lccode}, \cs{uccode}]
%\gr{8-bit number} (0-255); lowercase and
%uppercase codes \rdash these were treated above.
%\end{description}
The code names are as follows:
\begin{description}\item [\cs{catcode}]
\gr{4-bit number} (0--15); the category to which a character belongs.
This is treated in Chapter~\ref{mouth}.
\item [\cs{mathcode}]
\gr{15-bit number} (0--\verb-"7FFF-) or \verb-"8000-;
determines how a character is treated
in math mode. See Chapter~\ref{mathchar}.
\item [\cs{delcode}]
\gr{27-bit number} (0--\n{\hex7$\,$FFF$\,$FFF});
determines how a character is treated after
\cs{left} or \cs{right} in math mode.
See page~\pageref{delcodes}.
\item [\cs{sfcode}]
integer; determines how spacing is affected after this character.
See Chapter~\ref{space}.
\item [\cs{lccode}, \cs{uccode}]
\gr{8-bit number} (0-255); lowercase and
uppercase codes \rdash these were treated above.
\end{description}

%%\point Converting tokens into character strings
%\section{Converting tokens into character strings}
%\point Converting tokens into character strings
\section{Converting tokens into character strings}

%The command \cs{string} takes the next token and expands it
%\cstoidx string\par
%into a string of separate characters. Thus
%\begin{verbatim}
%\tt\string\control
%\end{verbatim}
%will give \cs{control} in the
%output, and
%\begin{verbatim}
%\tt\string$
%\end{verbatim}
%will give~\verb-$-, but, noting that the string 
%operation comes after the tokenizing,
%\begin{verbatim}
%\tt\string%
%\end{verbatim}
%will {\em not\/} give~\verb$%$,
%because the comment
%sign is removed by \TeX's input processor.
%Therefore, this command will `string' the first token on the next line.
The command \cs{string} takes the next token and expands it
\cstoidx string\par
into a string of separate characters. Thus
\begin{verbatim}
\tt\string\control
\end{verbatim}
will give \cs{control} in the
output, and
\begin{verbatim}
\tt\string$
\end{verbatim}
will give~\verb-$-, but, noting that the string 
operation comes after the tokenizing,
\begin{verbatim}
\tt\string%
\end{verbatim}
will {\em not\/} give~\verb$%$,
because the comment
sign is removed by \TeX's input processor.
Therefore, this command will `string' the first token on the next line.

%The \cs{string} command is executed by the expansion processor, thus
%it is expanded unless explicitly inhibited (see Chapter~\ref{expand}).
The \cs{string} command is executed by the expansion processor, thus
it is expanded unless explicitly inhibited (see Chapter~\ref{expand}).

%%\spoint Output of control sequences
%\subsection{Output of control sequences}
%\spoint Output of control sequences
\subsection{Output of control sequences}

%In the above examples the typewriter font was selected, because
%\cstoidx escapechar\par
%the Computer Modern roman font does not have a backslash character.
%However,
%\TeX\ need not have used the backslash character to display
%a control sequence: it uses character number \cs{escapechar}.
%This same value is also used when a control sequence is
%output with \cs{write}, \cs{message}, or \cs{errmessage},
%and it is used in the output of \cs{show}, \cs{showthe} and \cs{meaning}.
%If \cs{escapechar} is negative or more than~255,
%the escape character is not
%output; the default value (set in \IniTeX) is~92, the number
%of the backslash character.
In the above examples the typewriter font was selected, because
\cstoidx escapechar\par
the Computer Modern roman font does not have a backslash character.
However,
\TeX\ need not have used the backslash character to display
a control sequence: it uses character number \cs{escapechar}.
This same value is also used when a control sequence is
output with \cs{write}, \cs{message}, or \cs{errmessage},
and it is used in the output of \cs{show}, \cs{showthe} and \cs{meaning}.
If \cs{escapechar} is negative or more than~255,
the escape character is not
output; the default value (set in \IniTeX) is~92, the number
of the backslash character.

%For use in a  \cs{write} statement the \cs{string} can 
%in some circumstances be
%replaced  by \cs{noexpand} (see page~\pageref{expand:write}).
For use in a  \cs{write} statement the \cs{string} can 
in some circumstances be
replaced  by \cs{noexpand} (see page~\pageref{expand:write}).

%%\spoint Category codes of a \cs{string}
%\subsection{Category codes of a \cs{string}}
%\spoint Category codes of a \cs{string}
\subsection{Category codes of a \cs{string}}

%The characters that are the result of a \cs{string} command have 
%category code~12\index{category!12}, except for any spaces in 
%a stringed control sequence;
%they have category code~10\index{category!10}. Since inside a control
%sequence there are no category codes, 
%any spaces resulting from \cs{string} are
%of necessity only space {\em characters}, that is,
%characters with code~32.
%However, \TeX's input processor converts
%all space tokens that have a character code other than~32
%into character tokens with character code~32, 
%so the chances are pretty slim that
%`funny spaces' wind up in control sequences.
The characters that are the result of a \cs{string} command have 
category code~12\index{category!12}, except for any spaces in 
a stringed control sequence;
they have category code~10\index{category!10}. Since inside a control
sequence there are no category codes, 
any spaces resulting from \cs{string} are
of necessity only space {\em characters}, that is,
characters with code~32.
However, \TeX's input processor converts
all space tokens that have a character code other than~32
into character tokens with character code~32, 
so the chances are pretty slim that
`funny spaces' wind up in control sequences.

%Other commands with the same behaviour with respect to 
%category codes as \cs{string}, are
%\cs{number},
%\cs{romannumeral}, \cs{jobname}, \cs{fontname}, \cs{meaning},
%and \cs{the}.
Other commands with the same behaviour with respect to 
category codes as \cs{string}, are
\cs{number},
\cs{romannumeral}, \cs{jobname}, \cs{fontname}, \cs{meaning},
and \cs{the}.


%\endofchapter
%%%%% end of input file [char]
\endofchapter
%%%% end of input file [char]

\end{document}
